---
title: "P8122 HW3"
author: "Sabrina Lin stl2137"
date: "11/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tableone)
```

# Part 1

```{r}
### read in data
salary_dat <- read_csv("/Users/SabrinaLin/Documents/Fall_2020_Causal_Inference/Homework/HW3/p8122_hw3_stl2137/hW3\ data.csv") %>% 
  mutate(
    treat = as.factor(treat),
    black = as.factor(black),
    hispan = as.factor(hispan),
    married = as.factor(married),
    nodegree = as.factor(nodegree)
  )
```

* data consists of 10 variables measured for each individual:

  * an indicator of treatment assignment (job training), `treat` 
  
  * age in years, `age`
  
  * education in years, `educ`
  
  * an indicator for African-American, `black` 
  
  * an indicator for Hispanic, `hispan`
  
  * an indicator for married, `married`
  
  * an indicator for high school degree, `nodegree`
  
  * income in 1974, `re74` 
  
  * income in 1975 `re75` 
  
  * income in 1978, `re78`

* The variable `treat` is the treatment and the variables `re78` is the outcome.

## Subpart 1
Write the DAG representing this observational study including all variables provided. Describe all the variables in the graph.


## Subpart 2
Evaluate covariate balance in this observational study. Show a table or a plot. Interpret the results.

```{r}
## Construct a table
vars <- c("age", "educ", "black", "hispan", "married", "nodegree", "re74", "re75", "re78")

tab_presub <- CreateTableOne(vars = vars, strata = "treat", data = salary_dat, test = FALSE)

print(tab_presub, smd = TRUE)
```

* Given that we would like the SMD to be less than 0.2 and having seen 0.25 as a common guideline for SMD in the literature, there are several variables that surpass this rule of thumb. The SMD (standardized mean difference) for the variable `black` is very large at 1.671, indicating that the covariate balance for this variable is not good. The variables `married` and `re74` also have relatively large SMDs (0.721 and 0.596 respectively), also indicating that the covariate balance for these variables is not great. 


